Philosophy of Distributional Semantics\\ \large Best Practices for the Application of Computational Methods to Questions of Meaning

### 2017-05-28 pds/Paper/paper_pds.tex
advanced some in second part

\paragraph{Contribution: Drawing on ML discussion for DS}
In the present paper we want to draw on the existing efforts made by experts in the fields of ethics, computer science, public policy and law to extend these principles and guidelines to the field of DS. Specifically, we will use the \emph{Principles for Accountable Algorithms and a Social Impact Statement for Algorithms} (PAASISA) issued by an organization called \emph{Fairness, Accountability and Transparency in Machine Learning} - a consortium of researches and experts from private and public organizations working on machine learning. 
The PAASISA comprise five guiding principles: responsibility, explainability, accuracy, auditibility and fairness \cite{principles}.
We will assess potential violation of techniques used in distributional semantics for each of these principles and, where applicable, issue guidelines to prevent their negative impact upon vulnerable groups.

We want to focus on aspects of discrimination and bias in the outcome because these issues are least explored. Debates concerning privacy etc. would seem to follow the same lines as the discussion to establish general purpose ethical principles for machine learning, while the issue of discrimination and bias has a special relation to DS.

\paragraph{Methodoology}
The method employed by the present contribution is one of scanning previous efforts in a similar vein, identifying useful parts and applying them to ethical questions surrounding distributional semantics. There will be no foundational efforts undertaken to sustain the validity of the principles thus transferred.  Their validity is presupposed. The analysis contained in this contribution merely serves to disucss the usefulness of the principles in particular domain, and assert it where due.

\paragraph{Paper Overview}
The paper is structured as follows: In \hyperlink{sec2}{Section 2} we first give a very brief introduction to the field of machine learning in general, before providing some background on the algorithmic techniques employed in distributional semantics. Especially for the latter we want to provide overview over societal domains where such technology is currently applied and could conceivably be applied in the future. \hyperlink{sec3}{Section 3} contains a discussion of the important ways in which DS, as a particular application of big data analysis, differs from other instances of ML. For the present paper these particularities are important insofar as they entail differences when it comes to its risks and accountabilities across the spectrum of ML. In \hyperlink{sec4}{Section 4}, then, we try to apply the \emph{Principles for Accountable Algorithms and a Social Impact Statement for Algorithms} by FAT/ML to the specific case of DS. \hyperlink{sec5}{Section 5} contains concluding remarks while \hyperlink{sec6}{Section 6} lists the references.
\section{Machine Learning, Natural Language Processing and Distributional Semantics}\hypertarget{sec3}{ }
\paragraph{What is ML?}
The term \emph{Machine Learning} is an umbrella term for a variety of statistical methods that are able to \emph{learn} from a series of examples presented to them. The the set of examples - often called the \emph{training data} - is not simply memorized but the machine learning algorithm extracts generalizable knowledge \cite{domingos2012few}.
 What the algorithm learns are patterns and relationships within the training data and the result of the learning process is a statistical description - or \emph{model} - of the training data \cite{fayyad2001digital}.
\paragraph{What is NLP?}
\paragraph{Relation: ML - NLP}
\paragraph{What is Distributional Semantics?}
Distributional semantics is a linguistic theory of meaning looking to represent lexical meaning of expressions as a function of the context in which they appear. It theoretically foundation is the so-called \emph{distributional hypothesis} which holds that two expressions are similar in meaning, if they occur in similar contexts \cite{harris1954distributional}.
Althought it was originally developed as a distributional theory - in the sense that ... - in the middle of the 20th century, it can be seen as following the footsteps of structural linguists, notably Bloomfield \cite{bloomfield}
More info on DS. use \cite{boleda2016formal}

The theoretical framework of distributional semantics has become widely used and further developed as the advent and success of computational methods have enabled linguist to build large-scale models of linguistic meaning using digitalised corpora of text.

\paragraph{Relation: NLP - DS}
DS is naturally concerned with a narrower topic: meaning.\\
Meaning is a notoriously difficult topic conceptually.\\
DS sits at the intersection of ethically touchy topics: capturing/reporting/describing bias leads to reenforcing bias. results can be highly sensitive to data selection. even if results are valid there are issues of over/under-representation (scottish accent),

>>>>

\paragraph{Contribution: Drawing on ML discussion for DS}
In the present paper we want to draw on the existing efforts made by experts in the fields of ethics, computer science, public policy and law to extend these principles and guidelines to the field of distributional semantics. Specifically, we will use the \emph{Principles for Accountable Algorithms and a Social Impact Statement for Algorithms} (PAASISA) issued by the organization \emph{Fairness, Accountability and Transparency in Machine Learning} (FATML) - a consortium of researches and experts from private and public organizations working on machine learning. 
The PAASISA comprise five guiding principles: responsibility, explainability, accuracy, auditibility and fairness \cite{principles}.
We will assess potential violation of techniques used in distributional semantics for each of these principles and, where applicable, issue guidelines to prevent their negative impact upon vulnerable groups.

Currently, we want to focus on aspects of discrimination and bias in the outcome of distributional semantics models for two reasons. For one, these are the least explored issues for distributional semantics. Many of the ideas, guidelines and principles concerning privacy, double use or economic impact are directly transferable from the general machine learning debate to any sub-field that similarly employs big data and algorithms, as does computational distributional semantics. Debates concerning privacy etc. can follow the same lines as the discussion to establish general purpose ethical principles for machine learning. The issue of discrimination and bias, however, has a special relation to distributional semantics because it occupies in a unique position of computational language research. The central concern of distributional semantics, as its name gives away, is the concept of meaning. As we will argue presently, this relates it in a special way to many sensitive topics in the general machine learning discussion. Put succinctly, it sits at the intersection of concerns pertaining to data selection, validity or questionability of computational methods employed, interpretation of results and usage thereof.

\paragraph{Methodoology}
The method employed by the present contribution is one of scanning previous efforts in a similar vein, identifying useful parts and applying them to ethical questions surrounding distributional semantics. There will be no foundational efforts undertaken to sustain the validity of the principles thus transferred.  Their validity is presupposed. The analysis contained in this contribution merely serves to discuss the usefulness of the principles in particular domain, and assert it where due.

\paragraph{Paper Overview}
The paper is structured as follows: In \hyperlink{sec2}{Section 2} we first give a very brief description of machine learning, natural language processing and distributional semantics as well as their relation to each other. Especially for distributional semantics we want to provide overview over societal domains where such technology is currently applied and could conceivably be applied in the future. \hyperlink{sec3}{Section 3} contains a discussion of the important ways in which DS, as a particular application of big data analysis, differs from other instances of ML. For the present paper these particularities are important insofar as they entail differences when it comes to its risks and accountabilities across the spectrum of ML. In \hyperlink{sec4}{Section 4}, then, we try to apply the \emph{Principles for Accountable Algorithms and a Social Impact Statement for Algorithms} by FAT/ML to the specific case of DS. \hyperlink{sec5}{Section 5} contains concluding remarks while \hyperlink{sec6}{Section 6} lists the references.
\section{Machine Learning, Natural Language Processing and Distributional Semantics}\hypertarget{sec3}{ }
\paragraph{What is ML?}
The term \emph{Machine Learning} is an umbrella term for a variety of statistical methods that are able to \emph{learn} from a series of examples presented to them. The the set of examples - often called the \emph{training data} - is not simply memorized but the machine learning algorithm extracts generalizable knowledge \cite{domingos2012few}.
 What the algorithm learns are patterns and relationships within the training data and the result of the learning process is a statistical description - or \emph{model} - of the training data \cite{fayyad2001digital}.
\paragraph{What is NLP?}
Natural language processing is a field of research that uses computational tools to study natural language text or speech. The knowledge thus obtained can be used to develop computer systems to serve a wide variety of purposes. Principal applications include translation between natural languages, artificial intelligence, user interfaces, like chat-bots, for example, summarization of text or speech recognition \cite{chowdhury2003natural}.
\paragraph{Relation: ML - NLP}
The area of natural language processing frequently serves itself of methods from machine learning, for example to process very large textual corpora. Indeed, parts of natural language processing can be thought of as particular applications of machine learning techniques to the domain of natural language.
\paragraph{What is Distributional Semantics?}
Distributional semantics is a linguistic theory of meaning looking to represent lexical meaning of expressions as a function of the context in which they appear. It theoretically foundation is the so-called \emph{distributional hypothesis} which holds that two expressions are similar in meaning, if they occur in similar contexts \cite{harris1954distributional}.
%Althought it was originally developed as a distributional theory - in the sense that ... - in the middle of the 20th century, it can be seen as following the footsteps of structural linguists, notably Bloomfield \cite{bloomfield}

The theory of distributional semantics has been used successfully in computational semantics, as it has become possible to obtain semantic representations immediately from natural language data . The theoretical framework of distributional semantics has become widely used and further developed as the advent and success of computational methods have enabled linguist to build large-scale models of linguistic meaning using digitalised corpora of text. 
Following the distributional hypothesis, the meaning of an expression is a function of the context in which it occurs, where context can be construed in different ways. Most typically, it is other words surrounding an expression in a text - the sentence in which it appears, for example. Context is not restricted to such simple analysis, however, as it can possibly contain more complex syntactic relations within a text, or even visual, textual or auditory information gathered from data.
\cite{boleda2016formal}
Distributional semantics, then, takes the context in which an expression occurs and abstracts away from the particular contexts in a large corpus of natural language data. The mathematical vehicle for representing these abstractions are commonly vectors, but there are also other algebraic objects possible, as matrices or tensors. The meaning of an expression, as a function of its context, is then represented by the distribution across the dimensions of the vector. A set of words thus represented builds a semantic space - mathematically, a vector space - wherein one can study the semantic relationship between the expressions because some semantic relationships between expressions are mirrored by geometric relationship in these spaces.

\paragraph{Relation: NLP - DS}
Distributional semantics is concerned with a specific topic in the study of natural language, namely, meaning. It could thus be seen as one of many sub-fields of natural language processing research. 
It is notoriously difficult to come to grips with the concept of meaning from philosophical standpoint.
DS sits at the intersection of ethically touchy topics: capturing/reporting/describing bias leads to reenforcing bias. results can be highly sensitive to data selection. even if results are valid there are issues of over/under-representation (scottish accent),

### 2017-05-28 pds/Paper/pds_references.tex
advanced some in second part

%DS Papers

>>>>

@misc{principles,
  title = {{Discrimination algorithms disparate impact} Kernel Description},
  howpublished = {\url{https://www.theatlantic.com/business/archive/2015/09/discrimination-algorithms-disparate-impact/403969/
}},
  note = {Accessed: 2017-05-22}
}
@PROCEEDINGS{proceedings,
  title = {Proc.\ o.t. First Workshop on Ethics in NLP
      (Valencia, Spain, April 2017)},
  booktitle = {Proc.\ o.t. First Workshop on Ethics in NLP
      (Valencia, Spain, April 2017)},
  volume = 1,
  series = {},
  publisher = {},
  address = {Valencia, Spain},
  year = {2017}
}

@inproceedings{leidner2017ethical,
  author =       {Jochen L. Leidner, Vassilis Plachouras},
  title =        {Ethical by Design: Ethics Best Practices for Natural Language Processing},
  crossref =     {proceedings},
  pages =        {9--18},
}

%DS Papers
@article{wagner2015s,
  title={It's a man's wikipedia? Assessing gender inequality in an online encyclopedia},
  author={Wagner, Claudia and Garcia, David and Jadidi, Mohsen and Strohmaier, Markus},
  journal={arXiv preprint arXiv:1501.06307},
  year={2015}
}
@article{harris1954distributional,
  title={Distributional structure},
  author={Harris, Zellig S},
  journal={Word},
  volume={10},
  number={2-3},
  pages={146--162},
  year={1954},
  publisher={Taylor \& Francis}
}
@article{boleda2016formal,
  title={Formal Distributional Semantics: Introduction to the Special Issue},
  author={Boleda, Gemma and Herbelot, Aur{\'e}lie},
  journal={Computational Linguistics},
  year={2016},
  publisher={MIT Press}
}
@article{chowdhury2003natural,
  title={Natural language processing},
  author={Chowdhury, Gobinda G},
  journal={Annual review of information science and technology},
  volume={37},
  number={1},
  pages={51--89},
  year={2003},
  publisher={Wiley Online Library}
}
